Boosted Regression Trees for Predictive Auto-Tuning

### 2011-12-13 paper/draft1.tex
Various edits

%XXX:  Machine Learning for Auto-tuning: A Data-Drive Approach...} ???
%\title{Boosted Regression Trees for Data-Driven Auto-Tuning}
\title{Machine Learning for Predictive Auto-Tuning}
% NP: I like this title !
%XXX:  Machine Learning for Auto-tuning: A Data-Drive Approach...} ???

@@

\begin{abstract}
Auto-tuning is a widely used and effective technique for optimizing a
parametrized GPU code template for a particular computation on particular
hardware.  Its drawback is that thorough or exhaustive auto-tuning requires
compiling many kernels and calling each one many times; this process is slow.
Furthermore, library abstraction boundaries provide operations such as image
filtering and matrix multiplication, which actually correspond to a large set
of potential problem configurations with a wide variety of memory access
patterns and computational bottlenecks.  How can we draw on data from previous
empirical auto-tuning of related problems on related hardware to make a
just-in-time implementation decision for a novel problem?  This paper presents
a machine learning approach to auto-tuning, in which features of (a) the
current hardware platform, (b) the kernel configuration and (c) the problem
instance are passed to a regression model (boosted regression trees) which
predicts how much faster this kernel will be than a reference baseline.
Combinatorial optimization strategies that would normally implement auto-tuning
by evaluating kernel configurations on the real hardware are orders of
magnitude faster when evaluating the surrogate regression model instead.  We
validate our approach using the filterbank correlation kernel described in
\citet{pinto+cox:2011gcg}, where we find that 0.1 seconds (XXX) of hill
climbing on the regression model ({\em predictive} auto-tuning) can achieve an
average of 95\% of the speed brought by 120 seconds of empirical auto-tuning
XXX.  Our approach is not specific to filterbank correlation, or even to GPU
kernel auto-tuning: the approach of using a non-linear regression model on top
of simple features applies to a variety of problem types, kernel types, and
platforms.  \end{abstract}

@@

Although the future lies in low-power multi-core hardware designs, the field
lacks meaningful consensus on exactly how the different subsystems (memory,
communication and computation) should be efficiently integrated, modeled and
programmed. These systems have exhibited varying degrees of memory hierarchy
and multi-threading complexity and, as a consequence, they have been
increasingly relying on flexible but low-level software-controlled cache
management and parallelism \citep{asanovic2006landscape} in order to better
control and understand the various trade-offs among performance, reliability,
energy efficiency, production costs, etc. This evolution has profoundly
altered the landscape of application development: programmers are now
facing a wide diversity of low-level architectural features to write
high-performance \emph{and} portable code.

@@

One of the strategies for addressing these challenges is to use automatic
methods called ``auto-tuning''. Two major auto-tuning approaches have emerged
in the extensive literature covering the subject (see surveys in
\citep{vuduc2001statistical, demmel2005self, vuduc2005oski, williams2008auto,
datta2008stencil, cavazos2008intelligent, li2009note, park2011evaluation}):
analytical model-driven optimization and empirical optimization
\citep{yotov2003comparison}.

@@

The empirical optimization approach, on the contrary, seeks to find the best
performing code configuration by automatically generating many versions of a
parametrized kernel and benchmarking them on the actual hardware (possibly at
runtime, when contextual information about the hardware and software stack is
the richest). This method directly optimizes the metric(s) of interest and not
surrogates. One of its main advantage is that it allows any metric to be
optimized without loss of generality. It is indeed possible to formulate the
problem as a multi-objective optimization and minimize both speed \emph{and}
power consumption \citep{rahman2011automated}, a feat that renders the
analytical model-driven approach even more intractable. Due to its flexibility,
empirical auto-tuning has been successfully applied to build high-performance
domain-specific libraries including dense linear algebra
\citep{clint2001automated, bilmes1997optimizing}, sparse linear algebra
\citep{vuduc2005oski}, signal processing \citep{frigo1998fftw}, sorting
\citep{li2004dynamically}, general stencil operations \citep{kamil2010auto}, etc.

Even though this ``black-box'' approach does not necessarily try to build an
explicit model per se, it could, if care is taken, help to greatly accelerate
our understanding of the complex interactions between the set of optimizations,
their parameters and the hardware architectural features (XXX: that's what
we are trying to do w/ ML-based, decision trees, etc.).

@@

The empirical approach is very sensible to the choice of instrumented
optimizations \emph{and} to the search method. The size of the search space is
so large that the current best empirical auto-tuners can only consider
highly-specialized functions with a limited set of code transformations and
compiler options, on very few input domains \citep{ganapathi2009case}. Although
searching for good code configurations in this highly-discontinuous space can
be made embarrassingly parallel, it remains a very difficult and prohibitively
expensive combinatorial optimization problem as many variants of the code must
generated, compiled, and benchmarked on specific input domains with meaningful
statistics (that may require multiple runs). Consequently, most proposed
methods prune the space with hard-coded heuristics that offer little
generalization guarantees. This has been the main drawback of the empirical
approach compared to the model-driven approach where good code configurations
can be directly derived from the analytical model.

@@

To alleviate this major weakness, it seems intuitive to combine the two
approaches by first constraining the search space with an analytical model and
then exploring the reduced space empirically \citep{chen2005combining,
li2009note}. Unfortunately, this hybrid approach is still bound by the quality
of the analytical model, which remains hard to build by hand.

@@

In this paper, we propose to \emph{learn the model}, empirically, (XXX:
we'll describe what this means in more details later in the paper, riiiighhht
???) instead of deriving it analytically. By learning the model, one can
hope to get ``the best of both approaches'' and capitalize on the large body of
literature from the machine learning community.

@@

\subsection{Case Study}
Current languages and software engineering practices have conspired to make it difficult
to provide high-performance libraries for general mathematical operations on GPU devices.
The dominant approach to implementing high-performance GPU codes has been auto-tuning,
which is fundamentally slow. Auto-tuning requires running a large number of
sub-compilations and representative computations to actually {\em measure} which
of several possible implementations is fastest. General-purpose libraries such
as CUBLAS and CUFFT XXX hide this overhead by auto-tuning offline and using
intimate knowledge of the mathematical problem being solved as well as the
kernel implementations to heuristically choose a good computation strategy quickly when a
user asks for e.g. a matrix product of particular arguments.
Designing and testing libraries of this form is time-consuming, difficult, and provides little
reusable infrastructure to assist in the implementation of different computations.
Libraries for stencil computations and sparse matrix operations must start from scratch.

>>>>

%XXX:  Machine Learning for Auto-tuning: A Data-Drive Approach...} ???
%\title{Boosted Regression Trees for Data-Driven Auto-Tuning}
\title{Boosted Regression Trees for Predictive Auto-Tuning}
% NP: I like this title !
%XXX:  Machine Learning for Auto-tuning: A Data-Drive Approach...} ???

@@

\begin{abstract}
Template-based auto-tuning techniques are widely used and effective for optimizing a
parametrized GPU code template for a particular computation on particular
hardware \cite{XXXX}.  A drawback of this approach is that exhaustive, or even thorough, auto-tuning requires
compiling many kernels and calling each one many times, and this process is slow.
Furthermore, library abstraction boundaries provide operations such as image
filtering and matrix multiplication, which actually correspond to a large set
of potential problem configurations with a wide variety of memory access
patterns and computational bottlenecks.  How can we draw on data from previous
empirical auto-tuning of related problems on related hardware to make a
just-in-time implementation decision for a novel problem?  This paper presents
a machine learning approach to auto-tuning, in which features of the
current hardware platform, the kernel configuration and the problem
instance are passed to a regression model (boosted regression trees) which
predicts how much faster this kernel will be than a reference baseline.
Combinatorial optimization strategies that would normally implement auto-tuning
by evaluating kernel configurations on the real hardware are orders of
magnitude faster when evaluating the surrogate regression model instead.  We
validate our approach using the filterbank correlation kernel described in
\citet{pinto+cox:2011gcg}, where we find that 0.1 seconds (XXX) of hill
climbing on the regression model ({\em predictive} auto-tuning) can achieve an
average of 95\% of the speed brought by 120 seconds of empirical auto-tuning
XXX.  Our approach is not specific to filterbank correlation, nor to GPU
kernel auto-tuning; the approach of using a non-linear regression model on top
of simple features applies to a variety of problem types, kernel types, and
platforms.  \end{abstract}

@@

Although the future lies in low-power multi-core hardware designs, the field
lacks  consensus on exactly how the different subsystems (memory,
communication and computation) should be efficiently integrated, modeled and
programmed. These systems have exhibited varying degrees of memory hierarchy
and multi-threading complexity and, as a consequence, they have been
increasingly relying on flexible but low-level software-controlled cache
management and parallelism \citep{asanovic2006landscape} in order to better
control and understand the various trade-offs among performance, reliability,
energy efficiency, production costs, etc. This evolution has profoundly
altered the landscape of application development: programmers are now
facing a wide diversity of low-level architectural features to write
high-performance \emph{and} portable code.

@@

One strategy for addressing these challenges is to use one of a variety of automatic
methods known collectively as ``auto-tuning''. Two major auto-tuning approaches have emerged
in the extensive literature covering the subject (see surveys in
\citep{vuduc2001statistical, demmel2005self, vuduc2005oski, williams2008auto,
datta2008stencil, cavazos2008intelligent, li2009note, park2011evaluation}):
analytical model-driven optimization and empirical optimization
\citep{yotov2003comparison}.

@@

The empirical optimization approach, on the contrary, seeks to find the best
performing code configuration by automatically generating many versions of a
parametrized kernel and benchmarking them on the actual hardware (possibly at
runtime, when contextual information about the hardware and software stack is
the richest). This method directly optimizes the metric(s) of interest and not
surrogates. One of its main advantage is that it allows any metric to be
optimized without loss of generality. It is indeed possible to formulate the
problem as a multi-objective optimization and minimize both speed \emph{and}
power consumption \citep{rahman2011automated}, a feat that renders the
analytical model-driven approach even more difficult. Due to its flexibility,
empirical auto-tuning has been successfully applied to build high-performance
domain-specific libraries including dense linear algebra
\citep{clint2001automated, bilmes1997optimizing}, sparse linear algebra
\citep{vuduc2005oski}, signal processing \citep{frigo1998fftw}, sorting
\citep{li2004dynamically}, general stencil operations \citep{kamil2010auto}, etc.

Even though such a ``black-box'' approach does not try to build an
explicit model per se, it could, if care were taken, help to greatly accelerate
our understanding of the complex interactions between the set of optimizations,
their parameters and the hardware architectural features (XXX: that's what
we are trying to do w/ ML-based, decision trees, etc.).

@@

% DDC: "sensible to"?  what is meant here? "Sensitive to?"
The empirical approach is very sensible to the choice of instrumented
optimizations \emph{and} to the search method. The size of the search space is
so large that the current best empirical auto-tuners can only consider
highly-specialized functions with a limited set of code transformations and
compiler options, on very few input domains \citep{ganapathi2009case}. Although
searching for good code configurations in this highly-discontinuous space can
be made embarrassingly parallel, it remains a very difficult and prohibitively
expensive combinatorial optimization problem as many variants of the code must
generated, compiled, and benchmarked on specific input domains with meaningful
statistics (that may require multiple runs). Consequently, most proposed
methods prune the space with hard-coded heuristics that offer little
generalization guarantees. This has been the main drawback of the empirical
approach compared to the model-driven approach where good code configurations
can be directly derived from the analytical model.

@@

To alleviate this major weakness, it is intuitively appealing to combine the two
approaches by first constraining the search space with an analytical model and
then exploring the reduced space empirically \citep{chen2005combining,
li2009note}. Unfortunately, this hybrid approach is still bound by the quality
of the analytical model, which remains hard to build by hand.

@@

In this paper, we propose to \emph{learn the model}, empirically, (XXX:
we'll describe what this means in more details later in the paper, riiiighhht
???) instead of deriving it analytically. By learning the model, one can
hope to achieve elements of the best of both approaches and capitalize on the large body of
literature from the machine learning community.

@@

\subsection{Case Study}
Current languages and software engineering practices have conspired to make it difficult
to provide high-performance libraries for general mathematical operations on GPU devices.
An important approach in implementing high-performance GPU codes has been auto-tuning,
which is fundamentally slow. Auto-tuning requires running a large number of
sub-compilations and representative computations to {\em measure} which
of several possible implementations is fastest. General-purpose libraries such
as CUBLAS and CUFFT XXX hide this overhead by auto-tuning offline and using
intimate knowledge of the mathematical problem being solved as well as the
kernel implementations to heuristically choose a good computation strategy quickly when a
user asks for e.g. a matrix product of particular arguments.
Designing and testing libraries of this form is time-consuming, difficult, and provides little
reusable infrastructure to assist in the implementation of different computations.
For instance, even if one has already developed an auto-tuned library for dense matrix algebra, 
one must basically start from scratch to develop an auto-tuned library for
stencil computations or sparse matrix operations.
